The curious world of constructive mathematics

### 2023-09-14 algar-lecture-notes.tex
typo

The base step~$a = 0$ is trivial, since if~$a = 0$ then also~$a = 0 \vee a \neq
0$ (we can always weaken statements by addition additional disjuncts).

@@

As from any constructive proof, a computer program witnessing the asserted fact
can be extracted. It will (roughly) have type $\mathsf{Nat} \times \mathsf{Nat}
\to \mathsf{Bool}$, computing~$\mathsf{true}$ or~$\mathsf{false}$ depending on
whether its input agree or are distinct. Because the presented proof proceeded
by induction, the resulting program will proceed by recursion.

@@

An example of a result of type~(b) us the basic observation that every
inhabited set of natural numbers contains a minimal element. This result is not
available in constructive mathematics as stated, but there are two constructive
substitutes.

\begin{defn}A set~$X$ is \emph{inhabited} if and only if there exists an
element of~$X$. (Constructively, this is stronger than~$X$ not being
empty.)\end{defn}

@@

This lecture illustrates the latent constructive nature of classical
proofs with examples and presents two general metatheorems which
elucidate proof mining for constructive content. The author learned this
material from diverse texts and slides of Thierry Coquand; the
reference~\cite{coquand:classical} is a good start.
\end{intro}

@@

\begin{enumerate}
\item \emph{Eliminating the axiom of choice by the~$L$-translation.} Can the
axiom of choice ever help in proving arithmetical statements, those first-order
statements in which all quantifiers range over the natural numbers? Well, it
might. But a result of Gödel states that~\textsc{zfc} (Zermelo--Fraenkel set
theory with the axiom of choice) is conservative over~\textsc{zf} (ZF set
theory without it) for such statements -- hence all appeals to this axiom can
be mechanically eliminated from a given proof. This is true even if the proof
transcends the arithmetical realm and includes statements which are not
arithmetical; only the asserted claim is required to be arithmetical. Hence we
are free to use the axiom of choice, tranquil in knowing that we could always
reformulate our proofs without it.\footnote{Zermelo--Fraenkel set theory with
the axiom of choice is the go-to foundation of mathematics often cited as
supporting ``almost all'' of current mathematics, one important exception being
some (definitions and) results in category theory dealing with large
structures~\cite{shulman:set-theory,feferman:set-foundations}. A fundamental
result due to Gödel is that the axiom of choice ``holds in~$L$'', the
\emph{constructible universe}, even if it might not hold in~$V$, the true
universe of all sets. More precisely, if~\textsc{zfc} shows some
statement~$\varphi$, then~\textsc{zf} shows its~$L$-relativized
version~$\varphi^L$, where all quantifiers have been restricted to range
over~$L$ instead of~$V$. The conservation result follows because the natural
numbers ``are absolute between~$V$
and~$L$''~\cite{goedel:ac-gch,schoenfield:predicativity}.}

\item \emph{Eliminating the principle of excluded middle by the double-negation
translation and its variants.} At the price of
slightly modifying the asserted claim, the principle of excluded middle
can always be mechanically eliminated from a given proof. This
elimination procedure is facilitated by the \emph{double-negation translation}
reviewed below. In some cases, a refined translation even allows to preserve the asserted
claim exactly; this technique is variously known as \emph{Friedman's trick},
\emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
theorem}. To cite a specific instance of this phenomenon, classical~\textsc{zf}
set theory is conservative over its intuitionistic cousin~\textsc{izf}
for~$\Pi^0_2$-statements (statements of the form~$\forall\ldots\forall\_
\exists\ldots\exists\_ \%$, where all quantifiers in the final~``$\%$''
are bounded).

>>>>

The base step~$a = 0$ is trivial, since if~$a = 0$ then also~$a = 0 \vee a \neq
0$ (we can always weaken statements by adding additional disjuncts).

@@

As from any constructive proof, a computer program witnessing the asserted fact
can be extracted. It will (roughly) have type $\mathsf{Nat} \times \mathsf{Nat}
\to \mathsf{Bool}$, computing~$\mathsf{true}$ or~$\mathsf{false}$ depending on
whether its two inputs agree or are distinct. Because the presented proof proceeded
by induction, the resulting program will proceed by recursion.

@@

An example of a result of type~(b) uses the basic observation that every
inhabited set of natural numbers contains a minimal element. This result is not
available in constructive mathematics as stated, but there are two constructive
substitutes.

\begin{defn}A set~$X$ is \emph{inhabited} if and only if there exists an
element of~$X$.\end{defn}

Constructively, the condition that a set~$X$ is inhabited is stronger than~$X$
merely not being empty. The latter is equivalent to the statement that it is
\notnot the case that~$X$ is inhabited.

@@

This lecture illustrates the latent constructive nature of classical
proofs with examples and presents two general metatheorems which
elucidate proof mining for constructive content. The author learned this
material mostly from diverse texts and slides of Thierry Coquand; the
reference~\cite{coquand:classical} is a good start.
\end{intro}

@@

\begin{enumerate}
\item \emph{Eliminating the axiom of choice by the~$L$-translation.} Can the
axiom of choice ever help in proving arithmetical statements, those first-order
statements in which all quantifiers range over the natural numbers? Well, it surely
might. But a result of Gödel states that~\textsc{zfc} (Zermelo--Fraenkel set
theory with the axiom of choice) is conservative over~\textsc{zf} (ZF set
theory without it) for such statements -- hence all appeals to this axiom can
be mechanically eliminated from a given proof. This is true even if the proof
transcends the arithmetical realm and includes statements which are not
arithmetical; only the asserted claim is required to be arithmetical. Hence we
are free to use the axiom of choice, tranquil in knowing that we could always
reformulate our proofs without it.\footnote{Zermelo--Fraenkel set theory with
the axiom of choice is the go-to foundation of mathematics often cited as
supporting ``almost all'' of current mathematics, one important exception being
some (definitions and) results in category theory dealing with large
structures~\cite{shulman:set-theory,feferman:set-foundations}. A fundamental
result due to Gödel is that the axiom of choice ``holds in~$L$'', the
\emph{constructible universe}, even if it might not hold in~$V$, the true
universe of all sets. More precisely, if~\textsc{zfc} shows some
statement~$\varphi$, then~\textsc{zf} shows its~$L$-relativized
version~$\varphi^L$, where all quantifiers have been restricted to range
over~$L$ instead of~$V$. The conservation result follows because the natural
numbers ``are absolute between~$V$
and~$L$''~\cite{goedel:ac-gch,schoenfield:predicativity}.}

\item \emph{Eliminating the principle of excluded middle by the double-negation
translation and its variants.} At the price of
slightly modifying the asserted claim, the principle of excluded middle
can always be mechanically eliminated from a given proof. This
elimination procedure is facilitated by the \emph{double-negation translation}
reviewed below. In some cases, a refined translation even allows us to preserve the asserted
claim exactly; this technique is variously known as \emph{Friedman's trick},
\emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
theorem}. To cite a specific instance of this phenomenon, classical~\textsc{zf}
set theory is conservative over its intuitionistic cousin~\textsc{izf}
for~$\Pi^0_2$-statements (statements of the form~$\forall\ldots\forall\_
\exists\ldots\exists\_ \%$, where all quantifiers in the final~``$\%$''
are bounded).
